Musical elements in the discrete-time representation of sound

### 2017-07-07 article/acm/foobody.tex
added file ffor comparison

% notas:
% campo harmônico => harmonic field?
% anti-relative => counter-relative? (discussão sobre harmonia funcional)
%    associada relative => associated relative? (ver a discussão para mais...)
% medianas cromáticas => cromatic mediant
% tonalidade => key
% homônica => homonymous

\documentclass[
 aip,
 jmp,
 amsmath,amssymb,
 reprint,
]{revtex4-1}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{dcolumn}
\usepackage{bm}
\usepackage{multirow}
\usepackage{color}
\usepackage{xcolor}
\usepackage{enumerate}
\usepackage{float}
\usepackage{hyperref}

\newcommand{\massa}{{\large \textsc{massa}}}
\newcommand{\mass}{{\large \textsc{mass}}}
\newcommand{\figgus}{{\large \textsc{figgus}}}

\floatstyle{plain}
\newfloat{twocolequfloat}{t}{zzz}
\floatname{twocolequfloat}{Equation}

\begin{document}
\preprint{AIP/123-QED}

\title{Psychophysics of musical elements in the discrete-time representation of sound}

\author{Renato Fabbri}
 \email{renato.fabbri@gmail.com}
  \affiliation{ 
Instituto de F\'isica de S\~ao Carlos, Universidade de S\~ao Paulo (IFSC/USP)
}

\author{Vilson Vieira da Silva Junior}
  \homepage{http://automata.cc/}
  \email{vilson@void.cc}
 \altaffiliation[Also at ]{IFSC-USP}

\author{Ant\^onio Carlos Silvano Pessotti}
  \email{antoniopessotti@gmail.com}
 \altaffiliation[Instituto de Estudos da Linguagem ]{IEL/UNICAMP}

\author{D\'ebora Cristina Corr\^ea}
  \email{debcris.cor@gmail.com}
 \altaffiliation[Instituto de Ci\^encias Matem\'aticas e de Computa\c{c}\~ao ]{ICMC/USP}

%\author{Luciano da Fontoura Costa}
%  \homepage{http://cyvision.ifsc.usp.br/~luciano/}
%  \email{ldfcosta@gmail.com}
% \altaffiliation[Also at ]{IFSC-USP}
%
\author{Osvaldo N. Oliveira Jr.}
\homepage{www.polimeros.ifsc.usp.br/professors/professor.php?id=4}
\email{chu@ifsc.usp.br}
\altaffiliation[Also at ]{IFSC-USP}

\date{\today}
\begin{abstract}

    %The representation of the basic elements of music - such as notes, ornaments and interval structures - in terms of discrete audio signal is often used in software for music creation and design. Nevertheless, there is no unified approach that relates these elements to the sound discrete samples. In this article, each musical element is described by equations in sonic time samples, which are all implemented in scripts within an open source software toolbox, referred to as \massa\ (Music and Audio in Sequences and Samples). The fundamental element, the musical note with duration, volume, pitch and timbre, is related quantitatively to the characteristics of the discrete-time signal. Internal variations, such as tremolos, vibratos and spectral fluctuations, are also considered, which enables the synthesis of notes inspired by real instruments and new sonorities. With this representation of notes, resources are provided for the generation of musical structures, such as rhythmic meter, pitch intervals and cycles. The efficacy of these physical descriptions of basic musical elements was confirmed by the synthesis of small musical pieces within each frame: basic notes, incremented notes and notes in music. It is possible to synthesize whole albums through collage of the scripts and parametrization. 
    %The sample-based analytical description, and the paradigm of open source implementation, enables scientific experiments in precise and trustful ways. In fact, \massa\ has already been employed by external users for diverse purposes. Among these, it is mentioned an acoustic effect recognized by diverse individuals in mailing lists but not found in literature, and uses related to art and education.

Notes, ornaments and intervals are examples of basic musical elements.
Their representation as discrete-time digital audio plays a central role in software for music creation and design.
Nevertheless, there is no systematic relation, in analytical terms, of these musical elements to the sonic samples.
Such a compendium enables scientific experiments in precise and trustful ways, among educational and artistic uses.
This paper presents a comprehensive description of music in digital audio, within an unified approach.
Musical elements, like pitch, duration and timbre are expressed by equations on sample level.
This quantitatively relates characteristics of the discrete-time signal to musical qualities.
Internal variations, e.g. tremolos, vibratos and spectral fluctuations, are also considered
to operate within a note. Moreover, the generation of musical structures such as rhythmic meter,
pitch intervals and cycles, are attained canonically with notes. 
The availability of these resources in scripts is provided in public domain within the \massa\ toolbox - Music and Audio in Sequences and Samples. Authors observe that the implementation of sample-domain analytical results as open source can encourage concise research.
As further illustrated in the paper, \massa\ has already been employed by users for diverse purposes, 
including acoustics experimentation, art and education. The efficacy of these physical descriptions 
was confirmed by the synthesis of small musical pieces.
As shown, it is possible to synthesize whole albums through collage of scripts and parametrization.

\end{abstract}
\pacs{*43.66.-x,43.66.+y,05.10.-a} % PACS, the Physics and Astronomy
\keywords{psychophysics, acoustics, statistics, signal processing, digital audio, music}
\maketitle

\section{\label{sec:level1}Introduction}

Music is commonly defined as the art made by sounds and silences, where sound corresponds to the longitudinal wave of mechanical pressure. The human hearing system perceives sounds within the frequency bandwidth between $20Hz$ and $20kHz$, with the actual limits depending on the person, climate conditions and the sonic characteristics themselves~\cite{Roederer}. Since the speed of sound is $\approx 343.2 m/s$, these limits imply wavelengths of $\frac{343.2}{20} = 17.16\,m$ and $\frac{343.2}{20000}=17.16\,mm$. Such perception involves stimuli in bones, stomach, ears, transfer functions of head and torso, and processing by the nervous system~\cite{Roederer}. Obviously, the ear is a dedicated organ for the appreciation of these waves, which decomposes them into their sinusoidal spectra and delivers to the nervous system. The sinusoidal components are crucial to musical phenomena, as one can perceive in the composition of sounds of musical interest (such as harmonic sounds and noises, discussed in sections~\ref{sec:discNote} and~\ref{sec:internalVar}), and sonic structures of musical interest (such as tunings and scales, in section~\ref{sec:notesMusic}).

The representation of sound is commonly referred to as audio, although these terms are often used without distinction. Audio expresses waves from the capture by microphones or from direct synthesis, although these sources are not neatly distinguishable as captured sounds are processed to generate new sonorities. Digital audio specified by protocols that facilitate file transferring and storage often implies a quality loss. Standard representation of digital audio, on the other hand, assures perfect reconstruction of the analog wave, within any convenient precision. This paradigm consists of representing the audio with equally spaced samples, of $\lambda_s$ durations, each specified by a fixed number of bits. This is the Pulse Code Modulation (PCM) representation of sound. A sound in PCM audio is characterized by a sampling frequency $f_s=\frac{1}{\lambda_s}$ (also called the sampling rate), which is the number of samples used for representing a second of sound; and by a bit depth, which is the number of bits used for representing the amplitude of each sample. Figure~\ref{fig:PCM} shows $25$ samples of a PCM audio with $4$ bits each. The fixed $2^4=16$ values for the amplitude of each sample, with the regular spacing $\lambda_s$, yields the quantization error or noise. This noise diminishes as the bit depth increases.

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{figures/pcm}
        \caption{Pulse Code Modulation (PCM) audio: an analogical signal is represented by 25 samples with 4 bits each.}
        \label{fig:PCM}
\end{figure*}

The Nyquist theorem~\cite{Openheim} states that the sampling frequency is twice the maximum frequency of the represented signal. Thus, for general musical purposes, it is necessary to have samples in a rate at least twice the highest frequency heard by humans, that is, $f_s \geq 2\times 20kHz = 40kHz$. This is the fundamental reason for the adoption of sampling frequencies such as $44.1kHz$ and $48kHz$, standards in Compact Disks (CD) and broadcast systems (radio and television), respectively.

Within this framework, musical notes can be characterized. The note still stands paradigmatic as the 'fundamental unit' of musical structures and, in practice, it can unfold into sounds that uphold other approaches\footnote{Music from the twentieth century enlarged this traditional comprehension of music. This occurred in the concert music, especially in the concrete, electronic and electroacoustic styles. In the 1990s, it became evident that popular music had also incorporated sounds without defined pitch and temporal organization, to name a few of simple characteristics.}. 
    Notes are also convenient for another reason: the average listener -- and considerable part of the specialists -- presupposes rhythmic and pitch organization (made explicit in section~\ref{sec:notesMusic}) as fundamental musical properties, and these are developed in traditional musical theory in terms of notes.

\subsection{Contributions and paper organization}

This article aims at representing musical structures and artifices by their discrete-time sonic characteristics. Results include mathematical relations, usually in terms of samples, and their direct computer program implementations. Despite the general interests involved, there are few books and computer implementations that tackle the subject. These mainly focus on computer implementations and ways to mimic traditional instruments, with scattered mathematical formalisms. A compilation of the works and their contributions is in the bibliography~\cite{dissertacao}. To the best of the author's knowledge, there is a lack of articles on the topic. Moreover, although current computer implementations use the analytical descriptions presented in this study in a implicit manner, it seems that there has been no concise and mathematical description of the processes implemented.

In order to address this concise description of musical elements and structures, in terms of the digitalized sound, the objectives of this paper are:

\begin{enumerate}

\item Present a concise set of relations among musical basic elements and sequences of PCM audio samples.

\item Introduce a framework of sound synthesis with control at sample level and with potential uses in psycho-acoustical experiments and high-fidelity synthesis.

\item Provide the accessibility of the developed framework. The analytic relations presented in this article are implemented as public domain scripts, i.e. small computer programs using accessible technologies for better distribution and validation. This constitute the \massa\ toolbox, available in an open Git repository~\cite{gitBook}. These scripts are written in Python and make use of external libraries, such as Numpy that performs calls to Fortran routines for better computational efficiency. Part of the scripts have been transcribed to JavaScript and native Python with readiness, what favors their use in Web browsers such as Firefox and Chromium~\cite{numpy, audiolab, tutpython, python}. Furthermore, these are all open technologies, that is, published using licenses that allow copy, distribution and use of any part for research and derivatives. This way, the work presented here is embedded in recommended practices for availability and eases co-authorship processes~\cite{Raymond,Lessig}.

\item To provide a didactic presentation of the content to favor its apprehension and usage. It is worthwhile to mention that this subject comprises diverse topics on signal processing, music, psycho-acoustics and programming.

The remaining parts of this work are organized as follows: section~\ref{sec:discNote} characterizes the basic musical note; section~\ref{sec:internalVar} further develops internals of the musical note; section~\ref{sec:notesMusic} tackles the organization of musical notes in higher levels of musical structure~\cite{Wisnick,Webern,Lerdahl,Cook,Lacerda}. As these descriptions embody topics such as psycho-acoustics, cultural traditions, formalisms and protocols, the text points to external complements as needed~\cite{Zamacois,Schoenberg,microsound}.

The next section is a minimum text in which musical elements are presented side-by-side with the discrete-time samples they result. In order to account for validation and sharing, implementations on computer code of each one of these relations are gathered in the \massa\ toolbox together with little musical montage resulting from them.

%Representing musical structures and artifices by it's discrete sound characteristics is the purpose of this work. The results include mathematical relations and it's computer program implementations. Next section exposes the theoretical description, which is implemented as scripts in a one-to-one relation to the equations.

%\subsection{Sound and digital audio}

%Sound is a longitudinal wave of mechanical pressure. The frequency bandwidth between $20Hz$ and $20kHz$ is appreciated by human hearing system with boundaries dependent on the person, climate conditions and sonic characteristics itself~\cite{Roederer}. If considered the speed of sound of $\approx 343.2 m/s$, this limits corresponds to $\frac{343.2}{20} = 17.16\,m$ and $\frac{343.2}{20000}=17.16\,mm$.

%Human perception of sound involves captivation by bones, stomach, ears, transfer functions of head and processing torso and nervous system. Besides that, the ear is a dedicated organ to the capture of this waves. Its mechanism decomposes sound into its sinusoidal spectrum and delivers them to the nervous system. This sinusoidal components are crucial to musical phenomena, as one can observe in the composition of sounds with musical interest and in tunings and scales. Subsection~\ref{subsec:dicNote} exposes the presence of sinusoid in discrete-time audio and characterizes a basic musical note.

%The representation of sound is called audio (although these terms are often used without distinction), and this can be provenient from caption by microphones or from synthesis. Often enough, digital audio is specified by protocols that eases file storage and transferring, in cost of a direct representation or even some loss in quality. Standard representation of digital audio, on the other hand, consists of samples equally spaced by $\lambda_s$ durations in time, with each sample specified by a sample number of bits. This is called the Pulse Code Modulation representation of sound (PCM). A PCM digital sound is characterized by it's sampling frequency $f_s=\frac{1}{\lambda_s}$, also called sampling rate, and bit depth, which is the number of bits used of representing the amplitude of each sample. Figure~\ref{fig:PCM} shows $25$ samples of a PCM audio with $4$ bits each. The $2^4=16$ possible steps for each sample, together with the regular spacing $\lambda_s$ between them, introduces a quantization error. This noise, caused by this errors, diminishes as these spacing diminishes.

%\begin{figure*}
%    \centering
%        \includegraphics[width=\textwidth]{pcm}
%        \caption{Pulse Code Modulation (PCM) audio: an analogical signal is %represented by 25 samples with 4 bits each.}
%        \label{fig:PCM}
%\end{figure*}

%By the Nyquist theorem, it is known that half the sampling frequency is the maximum frequency of the signal. Thus, it is necessary to have a sampling frequency at least twice the highest frequency heard by humans $f_s \geq 2\times 20kHz = 40kHz$. This is the basis for the use of the sampling frequencies $44.1kHz$ and $48kHz$, standards in Compact Disks (CD) and Broadcast systems (Radio and TV), respectively.

%\subsection{Sonic art and musical theory}

%A common definition for music is the art made by sounds and silences. For the average listener -- and a reasonable part of specialists -- the notion of music presupposes rhythmic and pitch organization such as explained in subsection~\ref{subsec:notesMusic}. Music from the twentieth century enlarged this traditional comprehension of music. This occurred in concert music, specially in the concrete, electronic and electroacoustic styles. On the last decade of the century, it was evident that popular music has also incorporated sounds without defined pitch and temporal organization out of simple metrics. Even though, the note stands paradigmatic as a 'fundamental unit' of musical structures and, in practice, it can unfold in sounds that observe this recent developments. The definition and expansion of the musical note as the fundamental unit of music is approached in subsections~\ref{subsec:discNote} and~\ref{subsec:internalVar}, respectively. Subsection~\ref{subsec:notesMusic} tackles the organization of musical notes in a higher level~\cite{Wisnick,Webern,Lerdhal,Cook,Lacerda}.

%Musical theory embody topics as diverse as psycho-acoustics, cultural manifestations and formalisms. The section~\ref{sec:results} point this topics as needed and designate external complements~\cite{Zamacois,Schoenberg,microsound}.

%\subsection{Computational implementation}

%The results presented in this article are implemented as scripts, i.e.\ small computer programs implemented using accessible technologies for better distribution and validation. This constitute the \massa\ toolbox, available in public domain in an open Git repository~\cite{gitBook}. This scripts are written in Python and make use of external libraries Numpy and Scikits/Audiolab that performs calls to Fortran routines for better computational efficiency. Part of this code has been transcribed to JavaScript and native Python with readiness, what points to uses of this contribution in Web browsers such as Firefox and Chromium~\cite{numpy, audiolab, tutpython, python}.

%This are all open technologies, that is, published using licenses that allows copy, distribution and use of any part for research and derivatives. This way, the work here presented is available and eases co-authorship processes~\cite{Raymond,Lessig}.

%\subsection{Objectives}
%\label{subsec:objectives}
%The main goal of this article is to present a concise set of relations among musical basic elements and sequences of PCM audio samples. The next section is a minimum text in which music elements are presented side-by-side with the discrete-time samples they result. As validation and sharing, implementations on computer code of these relations and little musical pieces where gathered the \massa\ toolbox, available online.

%Secondary objectives include presenting a framework of sound synthesis with control at sample level, with potential uses in psychoacoustical experiments and high-fidelity synthesis. The didactic presentation of the content favors use and apprehension on a problem which calls diverse topics to be tackled: signal processing, music and psycho-acoustics, to name just a few.

%\subsection{Related work}
%Due to the general interest, and number of knowledge areas involved, there is a number of books and computer implementations that are of interest or present similarities to what is presented in this work. A more detailed comparison of them is pointed out in the bibliography~\cite{dissertacao}. There is almost no articles which could be found on the topic. In summary, there are computer implementations that use this analytical descriptions implicitly, but there is no such a concise and mathematical description of the processes implemented, as they aim to be libraries for sound and music. There are books on the topic that cover various aspects of effects and physical modeling, but none of them carry a concise description of musical elements and structures, but focus on aspects of musical sounds and ways to mimic traditional instruments.

\section{Characterization of the discrete-time musical note} \label{sec:discNote}\label{sec:notaDisc}

In diverse artistic and theoretical contexts, music is conceived as comprising fundamental units referred to as notes, ``atoms'' that constitute music itself~\cite{Wisnick, Lovelock, Webern}.
These notes are now understood as central elements of certain musical paradigms. In a cognitive perspective, the notes are seen as discretized elements that facilitate and enrich the flow of information through music~\cite{Roederer, Lacerda}.
Canonically, the principal properties of a musical note are duration, volume, pitch and timbre~\cite{Lacerda}, which can be quantified using the evenly time spaced sonic samples.

@@

\subsection{Duration}

The sample frequency $f_s$ is defined as the number of samples in each second of the discrete-time signal. Let $T_i=\{t_i\}$ be an ordered set of real samples separated by $\delta_s=1/f_s$ seconds. A musical note of duration $\Delta$ seconds is presented as a sequence $T_i^{\Delta}$ of $\lfloor \Delta . f_s \rfloor $ samples. That is, the integer part of the multiplication is considered, and an error of at most $\delta_s$ missing seconds is admitted, which is usually fine for musical purposes. As an example, if $f_s=44.1kHz \;\;\Rightarrow\;\;\lambda_s=\frac{1}{44100}\approx 23$ microseconds. Thus:

@@

\subsection{Volume}\label{subsec:volume}

The sensation of sound volume depends on reverberation and harmonic distribution, among other characteristics described in section~\ref{sec:varInternas}. One can get volume variations through the power of the wave~\cite{Chowning}:

@@

The quantity $V_{dB}$ has the decibel unit ($dB$). 
For each $10dB$ it is associated a ``doubled volume''.
A handy reference is $10dB$ for each step in the intensity scale: \emph{pianissimo}, \emph{piano}, \emph{mezzoforte}, \emph{forte} e \emph{fortissimo}. Other useful references are $dB$ values related to double amplitude or potency:

@@

An amplitude increase by a factor slightly above of 3 is required for yielding a doubled volume. These values are guides for increasing or decreasing the absolute values in the sample sequences with musical purposes. The conversion from decibels to amplitude gain (or attenuation) is straightforward:

@@

\subsection{Pitch}

To a note corresponds a sequence $T_i$ in which duration and volume are directly related to the size of the sequence and the amplitude of its samples, respectively. The pitch is specified by the fundamental frequency $f_0$ whose cycle has duration $\delta_{f_0}=1/f_0$. This duration multiplied by the sampling frequency $f_s$ results in the number of samples per cycle: $\lambda_{f_0}=f_s . \delta_{f_0} =f_a/f_0$.

@@

\subsection{Timbre}

In a sound with harmonic spectrum, the (wave) period corresponds to the cycle duration, given by the inverse of the fundamental frequency. The trajectory of the wave inside the period - called the waveform - defines a harmonic spectrum and, thus, a timbre\footnote{Timbre is a subjective and complex characteristic. Physically, the timbre is multidimensional and given by the temporal dynamics of energy in the spectral components that are harmonic or noisy. In addition, the word timbre is used to designate different things: one same note can have (be produced with) different timbres, an instrument has different timbres, two instruments of the same family have, at the same time, the same timbre that blends them in the same family, and different timbres as they are different instruments. It is worth mentioning that timbre is not always manifested in spectral traces, since cultural or circumstantial aspects alter our perception of timbre}. Sonic spectra with minimum differences can result in timbres with crucial differences and, consequently, distinct timbres can be produced using different spectra\cite{Roederer}.

@@

The square wave preserves only odd harmonics falling at $-6dB/octave$:

\begin{equation}\label{square}
     Q^f_i=\left\{ q^f_i \right\}= \left\{
         \begin{array}{l l}
              1 & \quad \text{for } \; \; (i\,\%\lambda_f)   <  \lambda_f /2  \\
              -1 & \quad \text{otherwise}\\
         \end{array} \right.
\end{equation}

@@

Figure~\ref{fig:formasDeOnda} presents the waveforms described in equations  ~\ref{sinusoid}, ~\ref{sawTooth}, ~\ref{triangular} and ~\ref{square} for $\lambda_f=100$ (period of $100$ samples). If $f_s=44.1kHz$, the PCM standard in Compact Disks, the wave has fundamental frequency $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Herz$, around A4, just above the central "C", whatever the waveform is.

@@

The resulting sound has the spectrum of the original waveform. As a consequence of its repetition in an identical form, the spectrum is perfectly harmonic, without noise or variations typical of the natural phenomenon. This can be observed in Figure~\ref{fig:espectroOboe}, that shows the spectrum of the original oboe note and a note with same duration, whose samples consists of the repetition of cycle of Figure~\ref{fig:formasDeOnda}. Summing up, the natural spectrum exhibits variations in the frequencies of the harmonics, in their intensities and some noise, while the note made from the sampled period has a perfectly harmonic spectrum.

@@

\subsection{Spectrum at sampled sound}

These sinusoidal components in the discretized sound have some particularities. Considering a signal $T_i$ and its corresponding Fourier decomposition $\mathcal{F}\langle T_i\rangle=C_i=\{c_i\}_0^{\Lambda-1}$, the recomposition is the sum of the frequency components as time samples\footnote{The factor $\frac{1}{\Lambda}$ could be distributed among the Fourier transform and its reconstruction, as preferred.}:

@@

This discussion makes it clear the equivalence ~\ref{equivalenciasFreqs}, ~\ref{equivalenciasModulos} and ~\ref{equivalenciasFases}:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The basic note}\label{notaBasica}

Let $f$ be such that it divides $f_a$\footnote{As pointed before, this limitation simplifies the explanation without losing generality, and will be overcome in the next section.}. A sequence $T_i$ of sonic samples separated by $\delta_a=1/f_a$ describes a musical note with a frequency of $f$ Hertz and $\Delta$ seconds of duration if, and only if, it has the periodicity $\lambda_f=f_a/f$ and size $\Lambda=\lfloor f_a . \Delta \rfloor $:

@@

\subsection{Spatial localization and spatialization}\label{subsec:spac}

A musical note always has a spatial localization, even though this is not one of its four basic properties: the note source position at the ordinary physical space. The reverberation in the environment in which sound occurs is the matter of \emph{spatialization}. Both fields, spatialization and spatial localization, are widely valued by audiophiles and the music industry~\cite{floEsp}.

\subsubsection{Spatial localization}

It is understood that the perception of sound localization occurs in our nervous system by three pieces of information: the delay of incoming sound between both ears, the difference of sound intensity at each ear and the filtering performed by the human body, including its chest, head and ears~\cite{Roederer, hrtf, Heeger}.

@@

Where $\zeta$ is the distance between ears $\zeta$, known to be $\zeta \approx 21.5cm$ in an adult human. Straightforward calculations result in ITD:

@@

Smith III points that reasonable concert rooms have total reverberation time of $\approx 1.9$ seconds, and that the period of first reflections is around $0.1$ seconds. With these values, under the given conditions, there are perceived wave pulses which propagate for $652.08$ m ($83.79k$ samples in $f_a=44.1kHz$) before reaching the ear. In addition, sound reflections made after propagation for $34.32$ m ($4.41k$ samples in $f_a=44.1kHz$) have incidences less distinct by hearing. These first reflections are particularly important to spatial sensation. The first incidence is the direct sound, described by ITD and IID of equations~\ref{eq:dti} and~\ref{eq:dii}. Assuming that each one of the first reflections, before reaching the ear, will propagate at least $3-30m$, depending on the room dimensions, the separation between the first reflections is $8-90ms$ ($\approx 350-4000$ samples in $f_a=44.1kHz$). It is experimentally verifiable that the number of reflections increases with the square of $\approx k.n^2$. A discussion about the use of convolutions and filtering to favor implementation of these phenomena is provided in subsection~\ref{subsec:mus2}, particularly in the paragraphs about reverberation.

@@

Figure~\ref{fig:mixagem} illustrates this overlapping process of discretized sound waves, each with 100 samples. If $f_a=44.1kHz$, the frequencies of the sawtooth, square and sine wave are, respectively: $\frac{f_a}{100/2}=882Hz$, $\frac{f_a}{100/4}=1764Hz$ and $\frac{f_a}{100/5}=2205Hz$. The samples duration is very short $\frac{f_a=44.1kHz}{100} \approx 2 \text{ milliseconds}$. One can complete the sequence with zeroes to sum (mix) sequences with different sizes.

@@

The expression defining the variable $\gamma_i$ can be understood as $f_s$ being added to $i$ at each second.
If $i$ is divided by the sample frequency, $\frac{i}{f_a}$
is incremented in $1$ at each second. Multiplied by the period, it results in $i \frac{\widetilde{\Lambda}}{f_a}$, which covers the period in each second. Finally, with frequency $f$ it results in $i . f \frac{\widetilde{\Lambda}}{f_a}$ which completes $f$ periods $\widetilde{\Lambda}$ in $1$ second, i.e. the resulting sequence presents the fundamental frequency $f$.

@@

Vibrato is a periodic variation in pitch (frequency) and tremolo is a period variation in volume (intensity)\footnote{The jargon may be different in other contexts. For example, in piano, tremolo is a vibrato in the classification used here. The definitions given here are common in contexts regarding music theory and electronic music. In addition, they are based on a broader literature than the one used for a specific instrument, practice or musical tradition~\cite{Lacerda,Harmonia}.}
For the general case, vibrato is described as:

@@

                \item A perfect interval (unison, perfect forth, perfect fifth, perfect octave), or a major interval (major second M2,
major third M3, major sixth M6 or major seventh M7), increased
by one semitone results in an augmented interval (e.g.\
augmented third aug3 with five semitones). The augmented forth
is also called tritone (aug4 ~ tri ~ TT).

>>>>

\section{Characterization of the discrete-time musical note} \label{sec:discNote}\label{sec:notaDisc}
In diverse artistic and theoretical contexts, music is conceived as comprising fundamental units referred to as notes, ``atoms'' that constitute music itself~\cite{Wisnick, Lovelock, Webern}.
These notes are now understood as central elements of certain musical paradigms.
In a cognitive perspective, the notes are seen as discretized elements that facilitate and enrich the flow of information through music~\cite{Roederer, Lacerda}.
Canonically, the principal properties of a musical note are duration, volume, pitch and timbre~\cite{Lacerda}, which can be quantified using the evenly time spaced sonic samples.

@@

\subsection{Duration}
The sample frequency $f_s$ is defined as the number of samples in each second of the discrete-time signal. Let $T_i=\{t_i\}$ be an ordered set of real samples separated by $\delta_s=1/f_s$ seconds.
A musical note of duration $\Delta$ seconds is presented as a sequence $T_i^{\Delta}$ of $\lfloor \Delta . f_s \rfloor $ samples.
That is, the integer part of the multiplication is considered, and an error of at most $\delta_s$ missing seconds is admitted, which is usually fine for musical purposes. As an example, if $f_s=44.1kHz \;\;\Rightarrow\;\;\lambda_s=\frac{1}{44100}\approx 23$ microseconds. Thus:

@@

\subsection{Volume}\label{subsec:volume}
The sensation of sound volume depends on reverberation and harmonic distribution, among other characteristics described in section~\ref{sec:varInternas}. One can get volume variations through the power of the wave~\cite{Chowning}:

@@

The quantity $V_{dB}$ has the decibel unit ($dB$). 
For each $10dB$ a ``doubled volume'' is associated.
A handy reference is $10dB$ for each step in the intensity scale: \emph{pianissimo}, \emph{piano}, \emph{mezzoforte}, \emph{forte} and \emph{fortissimo}. Other useful references are $dB$ values related to double amplitude or potency:

@@

An amplitude increase by a factor slightly above 3 is required for yielding a doubled volume. These values are guides for increasing or decreasing the absolute values in the sample sequences with musical purposes. The conversion from decibels to amplitude gain (or attenuation) is straightforward:

@@

\subsection{Pitch}
To a note corresponds a sequence $T_i$ in which duration and volume are directly related to the size of the sequence and the amplitude of its samples, respectively. The pitch is specified by the fundamental frequency $f_0$ whose cycle has duration $\delta_{f_0}=1/f_0$. This duration multiplied by the sampling frequency $f_s$ results in the number of samples per cycle: $\lambda_{f_0}=f_s . \delta_{f_0} =f_a/f_0$.

@@

\subsection{Timbre}
In a sound with harmonic spectrum, the (wave) period corresponds to the cycle duration, given by the inverse of the fundamental frequency. The trajectory of the wave inside the period - called the waveform - defines a harmonic spectrum and, thus, a timbre\footnote{Timbre is a subjective and complex characteristic. Physically, the timbre is multidimensional and given by the temporal dynamics of energy in the spectral components that are harmonic or noisy. In addition, the word timbre is used to designate different things: one same note can have (be produced with) different timbres, an instrument has different timbres, two instruments of the same family have, at the same time, the same timbre that blends them in the same family, and different timbres as they are different instruments. It is worth mentioning that timbre is not always manifested in spectral traces, since cultural or circumstantial aspects alter our perception of timbre}. Sonic spectra with minimum differences can result in timbres with crucial differences and, consequently, distinct timbres can be produced using different spectra\cite{Roederer}.

@@

The square wave preserves only odd harmonics falling at $-6dB/octave$:
\begin{equation}\label{square}
     Q^f_i=\left\{ q^f_i \right\}= \left\{
         \begin{array}{l l}
              1 & \quad \text{for } \; \; (i\,\%\lambda_f)   <  \lambda_f /2  \\
              -1 & \quad \text{otherwise}\\
         \end{array} \right.
\end{equation}

@@

Figure~\ref{fig:formasDeOnda} presents the waveforms described in equations  ~\ref{sinusoid}, ~\ref{sawTooth}, ~\ref{triangular} and ~\ref{square} for $\lambda_f=100$ (period of $100$ samples). If $f_s=44.1kHz$, the PCM standard in Compact Disks, the wave has fundamental frequency $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Hz$, around A4, just above the central "C", whatever the waveform is.

@@

The resulting sound has the spectrum of the original waveform. As a consequence of its repetition in an identical form, the spectrum is perfectly harmonic, without noise or variations typical of the natural phenomenon. This can be observed in Figure~\ref{fig:espectroOboe}, which shows the spectrum of the original oboe note and a note with same duration, whose samples consists of the repetition of cycle of Figure~\ref{fig:formasDeOnda}. Summing up, the natural spectrum exhibits variations in the frequencies of the harmonics, in their intensities and some noise, while the note made from the sampled period has a perfectly harmonic spectrum.

@@

\subsection{Spectrum at sampled sound}
These sinusoidal components in the discretized sound have some particularities. Considering a signal $T_i$ and its corresponding Fourier decomposition $\mathcal{F}\langle T_i\rangle=C_i=\{c_i\}_0^{\Lambda-1}$, the recomposition is the sum of the frequency components as time samples\footnote{The factor $\frac{1}{\Lambda}$ could be distributed among the Fourier transform and its reconstruction, as preferred.}:

@@

This discussion makes it clear the equivalence in ~\ref{equivalenciasFreqs}, ~\ref{equivalenciasModulos} and ~\ref{equivalenciasFases}:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The basic note}\label{notaBasica}
Let $f$ be such that it divides $f_a$\footnote{As pointed before, this limitation simplifies the explanation without losing generality, and will be overcome in the next section.}. A sequence $T_i$ of sonic samples separated by $\delta_a=1/f_a$ describes a musical note with a frequency of $f$ Hertz and $\Delta$ seconds of duration if, and only if, it has the periodicity $\lambda_f=f_a/f$ and size $\Lambda=\lfloor f_a . \Delta \rfloor $:

@@

\subsection{Spatial localization and spatialization}\label{subsec:spac}
A musical note always has a spatial localization, even though this is not one of its four basic properties: the note source position at the ordinary physical space. The reverberation in the environment in which sound occurs is the matter of \emph{spatialization}. Both fields, spatialization and spatial localization, are widely valued by audiophiles and the music industry~\cite{floEsp}.

\subsubsection{Spatial localization}
It is understood that the perception of sound localization occurs in our nervous system by three pieces of information: the delay of incoming sound between both ears, the difference of sound intensity at each ear and the filtering performed by the human body, including its chest, head and ears~\cite{Roederer, hrtf, Heeger}.

@@

where $\zeta$ is the distance between ears $\zeta$, known to be $\zeta \approx 21.5cm$ in an adult human. Straightforward calculations result in ITD:

@@

Smith III states that reasonable concert rooms have total reverberation time of $\approx 1.9$ s, and that the period of first reflections is around $0.1$ s. With these values, under the given conditions, there are perceived wave pulses which propagate for $652.08$ m ($83.79k$ samples in $f_a=44.1kHz$) before reaching the ear. In addition, sound reflections made after propagation for $34.32$ m ($4.41k$ samples in $f_a=44.1kHz$) have incidences less distinct by hearing. These first reflections are particularly important to spatial sensation. The first incidence is the direct sound, described by ITD and IID of equations~\ref{eq:dti} and~\ref{eq:dii}. Assuming that each one of the first reflections, before reaching the ear, will propagate at least $3-30m$, depending on the room dimensions, the separation between the first reflections is $8-90ms$ ($\approx 350-4000$ samples in $f_a=44.1kHz$). It is experimentally verifiable that the number of reflections increases with the square of $\approx k.n^2$. A discussion about the use of convolutions and filtering to favor implementation of these phenomena is provided in subsection~\ref{subsec:mus2}, particularly in the paragraphs about reverberation.

@@

Figure~\ref{fig:mixagem} illustrates this overlapping process of discretized sound waves, each with 100 samples. If $f_a=44.1kHz$, the frequencies of the sawtooth, square and sine wave are, respectively: $\frac{f_a}{100/2}=882Hz$, $\frac{f_a}{100/4}=1764Hz$ and $\frac{f_a}{100/5}=2205Hz$. The samples duration is very short $\frac{f_a=44.1kHz}{100} \approx 2 \text{ ms}$. One can complete the sequence with zeroes to sum (mix) sequences with different sizes.

@@

The expression defining the variable $\gamma_i$ can be understood as $f_s$ being added to $i$ at each second.
If $i$ is divided by the sample frequency, $\frac{i}{f_a}$
is incremented in $1$ at each second. Multiplied by the period, it results in $i \frac{\widetilde{\Lambda}}{f_a}$, which covers the period in each second. Finally, with frequency $f$ it results in $i . f \frac{\widetilde{\Lambda}}{f_a}$ which completes $f$ periods $\widetilde{\Lambda}$ in $1$ s, i.e. the resulting sequence presents the fundamental frequency $f$.

@@

Vibrato is a periodic variation in pitch (frequency) and tremolo is a period variation in volume (intensity).\footnote{The jargon may be different in other contexts. For example, in piano, tremolo is a vibrato in the classification used here. The definitions given here are common in contexts regarding music theory and electronic music. In addition, they are based on a broader literature than the one used for a specific instrument, practice or musical tradition~\cite{Lacerda,Harmonia}.}
For the general case, vibrato is described as:

@@

                \item A perfect interval (unison, perfect fourth, perfect fifth, perfect octave), or a major interval (major second M2,
major third M3, major sixth M6 or major seventh M7), increased
by one semitone results in an augmented interval (e.g.\
augmented third aug3 with five semitones). The augmented fourth
is also called tritone (aug4 ~ tri ~ TT).
